Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes

### 2023-11-03 mop/draft.tex
Updates from Overleaf
some latex error fixes unnoticed by overleaf

\title{Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes}
\author{Kevin Tan, Edward L. Ionides}
\date{}

@@

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. \kevin{TODO} We provide a method for incorporating the MCAP methodology in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.

Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual}. The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.

@@

\textbf{Intuition:} Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation. Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter. The intuition is as follows. Consider a particle filter likelihood estimate where the system (i.e. \textbf{both the process and measurement models}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

@@

We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$ \kevin{INSERT RESULTS HERE}

\begin{prop}[Correctness of DOP-$\alpha$ (Algorithm \ref{alg:dop})]
    When $\theta$ is evaluated at $\phi$, 
\end{prop}

\subsection{Measurement Off-Policy}

@@

>>>>

\title{Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes}
\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
\date{}

@@

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. \kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval. 
\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}

Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual} that we call \texttt{pompy}. The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.

@@

\textbf{Intuition:} Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation. Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter. The intuition is as follows. Consider a particle filter likelihood estimate where the system (i.e. \textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

@@

We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$

\begin{prop}[Correctness of DOP-$\alpha$ (Algorithm \ref{alg:dop})]
    \label{prop:dop-correctness}
    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood.

    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter, and its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf},

    and when $\alpha=0$

\end{prop}

\subsubsection{Practical Considerations}

It is only ever practical to evaluate $\theta$ at $\phi$. This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate. Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, the naively 
\subsection{Measurement Off-Policy Particle Filters}

@@

There is a workaround. Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick. If we have the state transitions evolve according to $\theta,$ but the resampling according to $\phi,$

We therefore make the following assumption in this section only:
\begin{aspt}[Differentiable Simulator]
    $\text{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
\end{aspt}

\section{Numerical Experiments}

\citet{liu2023bridging} have established that the straight-through estimator can be thought of as a first-order approximation of the gradient.
\kevin{Benchmark DOP vs straight-through estimator/reinmax in MOP}
